ML Estimation of Matrix t Distributions with Expectation Maximization

### 2017-01-03 paper.tex
Worked on page 2 and 3.

\begin{equation}
\int_{S\succ 0}e^{-\textrm{tr}(SA)}|S|^{x}dS=\Gamma_{p}\left(\frac{1+p+2x}{2}\right)|A|^{\frac{1+p+2x}{2}},\label{eq:intmultgammadef}
\end{equation}
where the domain of integration of $S$ is the set of real symmetric
positive definite matrices with dimension $p$, $A$ is a $p\times p$
nonsingular matrix, and $\Gamma_{p}(\cdot)$ is the multivariate gamma
function, given in terms of the standard gamma function $\Gamma(\cdot)$
by
\begin{equation}
\Gamma_{p}(a)=\pi^{p(p-1)/4}\prod_{j=1}^{p}\Gamma\left(a+\frac{1-j}{2}\right).\label{eq:multgammadef}
\end{equation}
For notation convenience, the matrix determinant is represented as
$|\cdot|$. We also define the multivariate digamma function
%
\begin{equation} \psi_{p}(a)=\frac{\partial}{\partial
	a}\log\Gamma_{p}(a)=\sum_{j=1}^{p}\psi\left(a+\frac{1-j}{2}\right).
\end{equation}

@@

Following \citet{gupta1999matrix}, we begin by deriving the
probability density of the matrix--$t$ distribution starting from
a matrix normal $N_{p,m}(M,S,\Omega)$ distributed random variate $X$
with Wishart $W_{p}\left(\Sigma,\nu\right)$ distributed row-scale
parameter $S$ of appropriate degrees of freedom $\nu$ (independent
from $X$), and marginalizing over $S$. Note that the row scale
parameter $S$ is a real symmetric positive definite matrix. We repeat
the results here because, along the way, we will utilize several
results that are useful in the derivation of the EM formulas.

The $p\times m$ random matrix $X$ follows $N_{p,m}(M,S,\Omega)$ has
the probability density %
\begin{equation}
	\frac{1}{(2\pi)^{pm/2}(\det\Omega)^{p/2}(\det
	S)^{m/2}}\exp\left(-\frac{1}{2}\textrm{tr}\left[\Omega^{-1}(X-M)^{\textrm{t}}S^{-1}(X-M)\right]\right)\label{eq:ndistpdf}
\end{equation}
where $S$ is a $p\times p$ positive definite matrix (the covariance of
the rows of $X$), $\Omega$ is a $m\times m$ positive definite matrix
(the covariance of the columns of $X$) and $M$ is a $p\times m$ matrix
(the mean of $X$).

In turn, the $p\times p$ matrix $S$ follows a Wishart distribution
\begin{equation}
W_{p}\left(\Sigma,\nu\right)\label{eq:wdist}
\end{equation}
 if its PDF is given by:
\begin{equation}
\frac{1}{2^{\nu p/2}(\det\Sigma)^{\nu/2}\Gamma_{p}(\frac{\nu}{2})}(\det S)^{\frac{\nu-p-1}{2}}\exp\left(-\textrm{tr}(\Sigma^{-1}S)\right)\label{eq:wdistpdf}
\end{equation}
where $\Sigma$ is a $p\times p$ positive definite matrix, $\nu>p-1$
is the number of degrees of freedom and $\Gamma_{p}$ is the
multivariate gamma function defined in \ref{eq:multgammadef}. We will
show that the random matrix
\begin{equation}
T=(S^{-\frac{1}{2}})^{t}X+M\label{eq:change}
\end{equation}
where $S^{\frac{1}{2}}$ is the Cholesky square root of $S$, follows
a matrix--$t$ distribution with $n=\nu-p+1$. The Jacobian of this
transformation is 
\begin{equation}
(\det S)^{\frac{1}{2}m}.\label{eq:jac}
\end{equation}

The joint PDF of $T$ and $S$ is the product of Eq.~\ref{eq:ndistpdf},
Eq.~\ref{eq:wdistpdf} and Eq.~\ref{eq:jac}:
\begin{equation}
\begin{split}j(T,S,M,\Sigma,\Omega,n)\coloneqq\frac{\pi^{-\frac{mp}{2}}(\det\Sigma)^{\frac{n+p-1}{2}}(\det\Omega)^{\frac{n+m-2}{2}}(\det S)^{\frac{m+n-2}{2}}}{2^{p\frac{n+m+p-1}{2}}\Gamma_{p}(\frac{n+p-1}{2})}\times\\
\exp\left(-\frac{1}{2}\textrm{tr}S(\Sigma+(T-M)\Omega^{-1}(T-M)^{\textrm{t}})\right)
\end{split}
\label{eq:jointpdf}
\end{equation}
Finally we obtain the PDF of the matrix--$t$ distribution
Eq.~\ref{eq:matrixtpdf} by integrating over $S$ in the above
expression, using directly Eq.~\ref{eq:intmultgammadef} and slightly
rearranging the terms (in particular, noting that
$\det(\Sigma+(T-M)\Omega^{-1}(T-M)^{\textrm{t}})=\det\Sigma\det(I_{p\times
p}+\Sigma^{-1}(T-M)\Omega^{-1}(T-M)^{\textrm{t}})$).

>>>>

\begin{equation}
\int_{S\succ 0}e^{-\textrm{tr}(SA)}|S|^{x}dS=\Gamma_{p}\left(\frac{1+p+2x}{2}\right)|A|^{\frac{1+p+2x}{2}},\label{eq:intmultgammadef}
\end{equation}
where the domain of integration is the set of real symmetric
positive definite matrices with dimension $p$, $A$ is a $p\times p$
nonsingular matrix, and $\Gamma_{p}(\cdot)$ is the multivariate gamma
function, given in terms of the standard gamma function $\Gamma(\cdot)$
by
\begin{equation}
\Gamma_{p}(a)=\pi^{p(p-1)/4}\prod_{j=1}^{p}\Gamma\left(a+\frac{1-j}{2}\right).\label{eq:multgammadef}
\end{equation}
For notation convenience, the matrix determinant is represented as
$|\cdot|$. We also define the multivariate digamma function
%
\begin{equation} \psi_{p}(a)=\frac{\partial}{\partial
	a}\log\Gamma_{p}(a)=\sum_{j=1}^{p}\psi\left(a+\frac{1-j}{2}\right).
\end{equation}

@@

Following \citet{gupta1999matrix}, we begin by deriving the
probability density of the matrix--$t$ distribution starting from
a matrix normal distributed random variate $X\sim N_{p,m}(M,S,\Omega)$
with Wishart distributed row scale parameter $S \sim W_{p}\left(\Sigma,\nu\right)$ of appropriate degrees of freedom $\nu$ (independent from $X$), and marginalizing over $S$. Note that the row scale
parameter $S$ is a real symmetric positive definite matrix. We repeat the results here because, along the way, we will utilize several results that are useful in the derivation of the EM formulas.

The $p\times m$ random matrix $X \sim N_{p,m}(M,S,\Omega)$ has density
\begin{equation}
	\frac{1}{(2\pi)^{pm/2}|\Omega|^{p/2}|S|^{m/2}}\exp\left(-\frac{1}{2}\textrm{tr}\left[\Omega^{-1}(X-M)^{\textrm{t}}S^{-1}(X-M)\right]\right),\label{eq:ndistpdf}
\end{equation}
where $S$ and $\Omega$ are symmetric positive definite matrices of dimension $p$ and $m$, respectively, and $M$ (the expectation of $X$) is a $p\times m$ matrix. The matrix $S$ is proportional to the covariance of the rows of $X$ while the matrix $\Omega$ is proportional to the covariance of the columns of $X$. In turn the $p\times p$ matrix $S\sim W_{p}\left(\Sigma,\nu\right)$ has density
\begin{equation}
\frac{|S|^{\frac{\nu-p-1}{2}}}{2^{\nu p/2}|\Sigma|^{\nu/2}\Gamma_{p}(\frac{\nu}{2})}\exp\left(-\textrm{tr}\left[\Sigma^{-1}S\right]\right),\label{eq:wdistpdf}
\end{equation}
where $\Sigma$ is a $p\times p$ symmetric positive definite matrix, $\nu>p-1$
is the number of degrees of freedom, and $\Gamma_{p}$ is the
multivariate gamma function defined in Eq.~\eqref{eq:multgammadef}.

We will show that the matrix
\begin{equation}
T=(S^{-\frac{1}{2}})^{t}X+M\label{eq:change}
\end{equation}
follows a matrix--$t$ distribution with degrees of freedom $n=\nu-p+1$. Here $S^{\frac{1}{2}}$ is the lower Cholesky factor of $S$, i.e., $S=S^{\frac{1}{2}}(S^{\frac{1}{2}})^t$. The Jacobian determinant of the transformation \eqref{eq:change} is
\begin{equation}
\left|\frac{\partial T}{\partial X}\right|=|S|^{-\frac{m}{2}}.\label{eq:jac}
\end{equation}

Using Eq.~\eqref{eq:ndistpdf}, Eq.~\eqref{eq:wdistpdf} and Eq.~\eqref{eq:jac} we can see the joint PDF of $T$ and $S$ is given as
\begin{equation}
\begin{split}j(T,S,M,\Sigma,\Omega,n)\coloneqq\frac{\pi^{-\frac{mp}{2}}(\det\Sigma)^{\frac{n+p-1}{2}}(\det\Omega)^{\frac{n+m-2}{2}}(\det S)^{\frac{m+n-2}{2}}}{2^{p\frac{n+m+p-1}{2}}\Gamma_{p}(\frac{n+p-1}{2})}\times\\
\exp\left(-\frac{1}{2}\textrm{tr}S(\Sigma+(T-M)\Omega^{-1}(T-M)^{\textrm{t}})\right).
\end{split}
\label{eq:jointpdf}
\end{equation}
Finally we obtain the density \eqref{eq:matrixtpdf} of the matrix--$t$ distribution
by integrating over $S$ in the above expression, using Eq.~\eqref{eq:intmultgammadef} and slightly
rearranging the terms. In particular noting that
$\det(\Sigma+(T-M)\Omega^{-1}(T-M)^{\textrm{t}})=\det\Sigma\det(I_{p\times
p}+\Sigma^{-1}(T-M)\Omega^{-1}(T-M)^{\textrm{t}})$.
